---
title: "Genome scans to detect selection"
subtitle: "Genome scans"
author:
    - "Per Unneberg"
date: "`r format(Sys.time(), '%d %B, %Y')`"
institute: NBIS
from: markdown+emoji
format:
  revealjs:
    theme:
      - white
      - ../nbis.scss
    self-contained: true
    toc: true
    toc-depth: 1
    slide-level: 3
    slide-number: true
    preview-links: true
    chalkboard: false
    footer: Genome scans to detect selection
    logo: https://nbis.se/assets/img/logos/nbislogo-green.svg
    smaller: true
    highlight-style: gruvbox
    fig-height: 12
    fig-width: 10
    fig-align: center
    navigation-mode: vertical
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
execute:
  echo: true
  warning: false
  cache: true
  include: true
  autodep: true
  eval: true
  error: true
opts:
  knitr:
    code-fold: true
    tidy: true
    fig-format: svg
---

### Setup  {visibility="hidden" .unnumbered .unlisted}

```{r libs}
#| echo: false
#| eval: true
#| cache: false
library(knitr)
library(ggplot2)
library(viridisLite)
library(gganimate)
library(patchwork)
library(tidyr)
library(dplyr)
bw <- theme_bw(base_size=18) %+replace% theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
theme_set(bw)
knitr::knit_hooks$set(inline = function(x) {
                      prettyNum(x, big.mark=" ")
                  })
```
```{r }
#| label: utility-functions
#| echo: false
#| eval: true
#| cache: false
```


# Selection

### Topics

- zfst
- pcadapt [@luu_PcadaptPackagePerform_2017]
- allele frequence difference (AFD) scan [@berner_AlleleFrequencyDifference_2019]
- BayeScan [@foll_GenomeScanMethodIdentify_2008]

- genetic load / mutational meltdown
  - VEP [@mclaren_EnsemblVariantEffect_2016]
  - SIFT [@kumar_PredictingEffectsCoding_2009]
  - GERP [@davydov_IdentifyingHighFraction_2010]

- local adaptation [@fitzpatrick_EcologicalGenomicsMeets_2015]
- conservation genetics [@allendorf_GenomicsFutureConservation_2010] [@formenti_EraReferenceGenomes_2022a]


