---
title: "Linked selection"
author:
    - "Per Unneberg"
format:
  revealjs:
    footer: Linked selection
---

## Setup  {visibility="hidden" .unnumbered .unlisted}

{{< include ../_knitr.qmd >}}

{{< include ../_rlibs.qmd >}}

```{r }
#| echo: false
#| eval: true
#| cache: false
library(viridisLite)
library(gganimate)
library(patchwork)
```

# Selection

## Topics

- direct selection(?)

# Linked selection

## Tajima's D

## Selective sweep

Use [Figure 2 @charlesworth_EffectsOfSelection_2021]

## Selective sweep

<!-- markdownlint-disable MD013 -->

```{bash }
#| label: pgip-slim-sweep-command
#| echo: true
#| eval: false
#| code-fold: true
pgip-slim --seed 42 -n 1000 -r 1e-6 -m 1e-7 --threads 12 recipes/slim/selective_sweep.slim -l 1000000 --outdir results/slim
pgip-tsstat results/slim/slim*.trees -n 10 --seed 31 -s pi -s S -s TajD -w 500 --threads 10 | gzip -v - > results/slim/selective_sweep.w500.csv.gz
```

```{r }
#| label: pgip-load-slim-data
#| echo: false
#| eval: true
data <- tidyr::tibble(read.csv("../../data/results/slim/selective_sweep.w500.csv.gz", header = TRUE))
a <- sum(1 / seq(10 - 1))
data$thetaW <- data$S / a
data_mean <- as.data.frame(data) %>%
  select(windows, TajD, pi, S, thetaW) %>%
  group_by(windows) %>%
  summarise(TajD = mean(TajD, na.rm = TRUE), pi = mean(pi, na.rm = TRUE), S = mean(S, na.rm = TRUE), thetaW = mean(thetaW, na.rm = TRUE)) %>%
  mutate(fn = "mean", pi_minus_thetaW = pi - thetaW)
L <- 1e6
xi <- L / 2
```

<!-- markdownlint-enable MD013 -->

:::: {.columns}

::: {.column width="50%"}

<!-- markdownlint-disable MD013 -->

```{r }
#| label: pgip-slim-sweep-plot-pi-S
#| echo: false
#| eval: true
#| fig-align: center
#| fig-height: 5
#| fig-width: 10
x <- data_mean %>% pivot_longer(
  cols = c("pi", "thetaW", "pi_minus_thetaW"),
  names_transform = list(name = ~ readr::parse_factor(.x, levels = c("pi", "thetaW", "pi_minus_thetaW")))
)
ggplot(x, aes(windows, value, color = name)) +
  geom_segment(aes(x = xi, y = -1, xend = xi, yend = -2.5), color = "black", arrow = arrow(length = unit(0.4, "cm"), type = "closed")) +
  xlab("Chromosomal position") +
  ylab("diversity") +
  geom_line(linewidth = 1) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.title = element_blank()) +
  scale_y_continuous(expand = c(0, 0), limits = c(-2.5, 10)) + # , limits = c(-0.001, 0.0045)) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, L)) +
  scale_colour_manual(
    name = "name", values = c("blue", "red", "green"),
    labels = expression(pi, theta[W], pi - theta[W])
  )
```

<!-- markdownlint-enable MD013 -->

:::

::: {.column width="50%"}

<!-- markdownlint-disable MD013 -->

```{r }
#| label: pgip-slim-sweep-plot-tajd
#| echo: false
#| eval: true
#| fig-align: center
#| fig-height: 5
#| fig-width: 10
x <- subset(data, fn %in% levels(factor(data$fn))[1:50])
psweep_static <- ggplot(x, aes(windows, TajD, group = fn)) +
  geom_line(color = "lightgray", linewidth = 0.5) +
  xlab("Chromosomal position") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  geom_segment(aes(x = xi, y = -2, xend = xi, yend = -2.5), arrow = arrow(length = unit(0.4, "cm"), type = "closed")) +
  scale_y_continuous(expand = c(0, 0), limits = c(-2.5, 2.5)) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, L))
p <- psweep_static + transition_manual(fn, cumulative = TRUE)
p
```

<!-- markdownlint-enable MD013 -->

:::

::::

The effect of a selective sweep on different statistics.

::: {.notes }

N=1e4, mu=5e-8, pi=4*N*mu=0.004, seek S=a*pi=0.01, so a=S/pi which
gives n. Need to ramp up recombination rate to see the dip in TajD
clearly.

pgip-slim --seed 42 -n 1000 -r 1e-6 -m 5e-8 --threads 12
recipes/slim/selective_sweep.slim -l 1000000 --outdir results/slim

2022-11-23: Current results based on mu=1e-7

Q: not clear how Rasmus Nielsen scales statistics to have mean=1
[@nielsen_MolecularSignaturesNatural_2005]

:::

## Selective sweep

<!-- markdownlint-disable MD013 -->

```{bash }
#| label: pgip-slim-sweep-command-2
#| echo: true
#| eval: false
#| code-fold: true
pgip-slim --seed 42 -n 1000 -r 1e-6 -m 1e-7 --threads 12 recipes/slim/selective_sweep.slim -l 1000000 --outdir results/slim
pgip-tsstat results/slim/slim*.trees -n 10 --seed 31 -s pi -s S -s TajD -w 500 --threads 10 | gzip -v - > results/slim/selective_sweep.w500.csv.gz
```

<!-- markdownlint-enable MD013 -->

:::: {.columns}

::: {.column width="50%"}

<!-- markdownlint-disable MD013 -->

```{r }
#| label: pgip-slim-sweep-plot-pi-S-2
#| echo: false
#| eval: true
#| fig-align: center
#| fig-height: 5
#| fig-width: 10
x <- data_mean %>% pivot_longer(
  cols = c("pi", "thetaW", "pi_minus_thetaW"),
  names_transform = list(name = ~ readr::parse_factor(.x, levels = c("pi", "thetaW", "pi_minus_thetaW")))
)
ggplot(
  x, aes(windows, value, color = name)
) +
  geom_segment(aes(x = xi, y = -1, xend = xi, yend = -2.5), color = "black", arrow = arrow(length = unit(0.4, "cm"), type = "closed")) +
  xlab("Chromosomal position") +
  ylab("diversity") +
  geom_line(linewidth = 1) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.title = element_blank()) +
  scale_y_continuous(expand = c(0, 0), limits = c(-2.5, 10)) + # , limits = c(-0.001, 0.0045)) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, L)) +
  scale_colour_manual(
    name = "name", values = c("blue", "red", "green"),
    labels = expression(pi, theta[W], pi - theta[W])
  )
```

<!-- markdownlint-enable MD013 -->

:::

::: {.column width="50%"}

```{r}
#| label: pgip-slim-sweep-plot-tajd-2
#| echo: false
#| eval: true
#| fig-align: center
#| fig-height: 5
#| fig-width: 10
psweep_static + geom_line(data = data_mean, color = "red", linewidth = 1.5)
```

:::

::::

The effect of a selective sweep on different statistics.

## References
